---
title: "rCISSVAE Vignette"
author: "Dani V"
date: "2025-08-04"
lightbox: auto
format:
   html:
    toc: true
    code-fold: true
    code-copy: true
execute:
  echo: true
  warning: false
  error: false
  freeze: auto
  eval: true
  fig-dpi: 600
---

## Installing package

Install devtools or remotes if not already installed: 

```{r}
install.packages("remotes")
# or
install.packages("devtools")

```

The rCISSVAE package can be installed with:

```{r}
remotes::install_github("CISS-VAE/rCISS-VAE")
# or
devtools::install_github("CISS-VAE/rCISS-VAE")

```

## Ensuring correct virtual environment for reticulate

This package uses `reticulate` to interface with the python version of the package `ciss-vae`. 

Therefore, it is necessary to make sure that you have a venv or conda environment set up that has the `ciss-vae` package installed. 

If you are comfortable creating an environment and installing the package, great! Then all you need to do is tell reticulate where to point. 

**For Venv**
```{r}
reticulate::use_virtualenv("./.venv", required = TRUE)

```

**For conda**

```{r}
reticulate::use_condaenv("myenv", required = TRUE)
```

### Virtual environment helper function

If you do not want to manually create the virtual environment, a helper function is being developed. 

## Quickstart

Once reticulate is pointing to the virtual environment containing the `ciss_vae` python package, you can either use the `run_cissvae` function or the `autotune_cissvae` function. 

If you know what hyperparameters you want to use for the model, use the `run_cissvae` function. 

### Run Cissvae
Your data should be in a DataFrame format with optional index. If you already have clusters you want to use, they should be in a separate vector from the dataframe. If you do not have clusters to begin with, set 'clusters' in `run_cissvae()` to NULL. 

Optionally, you can set 'return_model' to TRUE to include the model as part of the return object. This allows you to use `plot_cissvae_arch()` to plot your model architecture. 

```{r}
library(tidyverse)
library(reticulate)
library(rcissvae)

reticulate::use_virtualenv("./.venv", required = TRUE)


df_missing = read_csv("ign_data/df_missing.csv") %>% rename( "index" = `...1`)
clusters = read_csv("ign_data/clusters.csv")

dat = run_cissvae(
  data = df_missing,
  index_col = "index",
  clusters = clusters$clusters,
  epochs = 500,
  return_silhouettes=TRUE,
  verbose=TRUE,
  return_model = TRUE,
  layer_order_enc = c("unshared", "shared", "unshared"),
  layer_order_dec = c("shared", "unshared", "shared")
)

## retrieve results
imputed_df = dat$imputed
silhouette = dat$slhouettes ## if return_silhouettes = TRUE

## optional
plot_cissvae_arch(dat$model, title = "CISSVAE model") ## if return_model == TRUE
```

## Autotune

If you want to perform hyperparameter tuning, you can use the `autotune_cissvae()` function. 

This function wraps the cluster dataset creation, search space creation and running the autotune into one function so there are a lot of parameters to be set. 

For the SearchSpace args, 
Types of parameters:  
- reticulate::tuple(min, max, step) -\> creates a range   
- \c(a, b, c\) -\> select value from list
- x -\> set param as non-tunable    


```{r}
library(tidyverse)
library(reticulate)
#library(rcissvae)

reticulate::use_virtualenv("./.venv", required = TRUE)

df_missing = read_csv("ign_data/df_missing.csv") %>% rename( "index" = `...1`)
clusters = read_csv("ign_data/clusters.csv")

aut = autotune_cissvae (
  data = df_missing,
  index_col              = "index",
  clusters = clusters$clusters,
  save_model_path        = NULL,
  save_search_space_path = NULL,
  n_trials               = 2, ## number of autotune trials to run
  study_name             = "vae_autotune", ## name of optuna study
  device_preference      = "cuda",
  show_progress          = FALSE, ## set TRUE to show progressbar
  optuna_dashboard_db    = NULL, ## optional sqlite3 database, eg: "sqlite:///db.sqlite3"
  load_if_exists         = TRUE, ## if TRUE, loads existing study of name 'study_name' if one exists
  seed                   = 42, 
  verbose                = FALSE, 

  ## SearchSpace args
  num_hidden_layers = c(1, 4),  
  hidden_dims       = c(64, 512),
  latent_dim        = c(10, 100),
  latent_shared     = c(TRUE, FALSE),
  output_shared     = c(TRUE, FALSE),
  lr                = reticulate::tuple(1e-4, 1e-3),
  decay_factor      = c(0.9, 0.999),
  beta              = 0.01,
  num_epochs        = 10, ## setting this low for demo purposes
  batch_size        = 4000,
  num_shared_encode = c(0, 1, 3),
  num_shared_decode = c(0, 1, 3),
  refit_patience    = 2,
  refit_loops       = 10, ## setting this low for demo purposes
  epochs_per_loop   = 10, ## setting this low for demo purposes
  reset_lr_refit    = c(TRUE, FALSE)
)

# Access results:
imputed   <- aut$imputed
best_model <- aut$model
study     <- aut$study
results   <- aut$results


## optional
plot_cissvae_arch(best_model, title = "Best model from Autotune")
```